plyranges: a grammar for transforming genomics data

### 2018-05-15 paper.tex
to generic style over plos - ready for biorxiv

% Template for PLoS
% Version 3.4 January 2017
\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
% \bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below
\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

\usepackage{forarray}
\usepackage{xstring}
\newcommand{\getIndex}[2]{
  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
}

@@

\newcommand{\getAff}[1]{
  \getIndex{#1}{Monash University,Genentech}
}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{plyranges: a grammar for transforming genomics data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
Stuart Lee\textsuperscript{\getAff{Monash University}},
Michael Lawrence\textsuperscript{\getAff{Genentech}},
Di Cook\textsuperscript{\getAff{Monash University}}\\
\bigskip
\textbf{\getAff{Monash University}}Department of Econometrics and Business Statistics, Clayton, Victoria,
Australia\\
\textbf{\getAff{Genentech}}Bioinformatics and Computational Biology, Genentech, Inc., South San
Francisco, California, United States of America\\
\bigskip
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The Bioconductor project has created many useful data abstractions for
analysing high-throughput genomics experiments. However, there is a
cognitive load placed on a user in learning a data abstraction and
understanding its appropriate use. Throughout a standard workflow, a
user must navigate and know many of these abstractions to perform an
genomic analysis task, when a single data abstraction, a GRanges object
will suffice. The GRanges class naturally represent genomic intervals
and their associated measurements. By recognising that the GRanges class
follows `tidy' data principles we have created a grammar of genomic data
transformation. The grammar defines verbs for performing actions on and
between genomic interval data. It provides a principled way of
performing common genomic data analysis tasks through a coherent
interface to existing Bioconductor infrastructure, resulting in human
readable analysis workflows. We have implemented this grammar as a
Bioconductor/R package called plyranges.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Several domain specific languages (DSLs) enable scientists to process
and reason about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the Genome Query
Language (GQL) and its distributed implementation GenAp which use an
SQL-like syntax for fast retrieval of information of unprocessed
sequencing data {[}1{]}; {[}2{]}. Similarly, the Genometric Query
Language (GMQL) implements a relational algebra for combining genomic
datasets {[}3{]}. The command line application BEDtools develops an
extensive algebra for performing arithmetic between two or more sets of
genomic regions {[}4{]}. All of the aforementioned DSLs are designed to
be evaluated either at the command line or embedded in scripts for batch
processing. They exist in a sparse ecosystem, mostly consisting of UNIX
and database tools that lack biological semantics and operate at the
level of files and database tables.

The Bioconductor/R packages \texttt{IRanges} and \texttt{GenomicRanges}
{[}5--7{]} define a DSL for analysing genomics data with R, an
interactive data analysis environment that encourages reproducibility
and provides high-level abstractions for manipulating, modelling and
plotting data, through state of the art methods in statistical
computing. The packages define object-oriented (OO) abstractions for
representing genomic data and enable interoperability by allowing users
and developers to use these abstractions in their own code and packages.
Other genomic DSLs that are embedded in programming languages include
pybedtools and valr {[}8,9{]}, however these packages lack the
interoperability provided by the aforementioned Bioconductor packages
and are not easily extended.

@@

Recently, there has been a concerted, community effort to standardize R
data structures and workflows around the notion of tidy data {[}10{]}. A
tidy dataset is defined as a tabular data structure that has
observations as rows and columns as variables, and all measurements
pertain to a single observational unit. The tidy data pattern is useful
because it allows us to see how the data relate to the design of an
experiment and the variables measured. The \texttt{dplyr} package
{[}11{]} defines an API that maps notions from the general relational
algebra to operations on tidy data. It expresses each operation as a
cohesive, endomorphic verb. Taken together these features enable a user
to write human readable analysis workflows.

@@

\hypertarget{genomic-relational-algebra}{%
\section{Genomic relational algebra}\label{genomic-relational-algebra}}

\hypertarget{data-model}{%
\subsection{Data model}\label{data-model}}

\begin{figure}
{\centering \includegraphics[width=400pt]{diagrams/GRanges}}
\caption{An illustration of the GRanges data model for a
sample from an RNA-seq experiment. The core components of the data model
inlcude a seqname column (representing the chromosome), a ranges column
which consists of start and end coordinates for a genomic region, and a
strand identifier (either positive, negative, or unstranded). Metadata
are included as columns to the right of the dotted line as annotations
(gene\_id) or range level covariates (GC score and count).}
\label{fig:GRanges} 
\end{figure}

The \texttt{plyranges} DSL is built on the core Bioconductor data
structure GRanges, which is capable of representing all types of genomic
data at a semantic level that roughly matches the intuition of most
users. A GRanges is essentially a table, with columns for the
chromosome, start and end coordinates, and the strand, along with an
arbitrary set of additional columns, consisting of measurements or
metadata specific to the data type or experiment (figure
\ref{fig:GRanges}).

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{4cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarise()}} & aggregate over column(s) \\ 
   Aggregation & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
   \hline
   &  \textbf{\emph{mutate()}} & modifies any column \\
   & \textbf{\emph{select()}} & select columns \\
  Arithmetic (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
   & \emph{stretch()} & extend range by fixed amount \\
   &  \emph{shift\_(direction)} & shift coordinates \\
   & \emph{flank\_(direction)} & generate flanking regions \\
   & \emph{\%intersection\% } & row-wise intersection \\
   & \emph{\%union\%} & row-wise union \\
   & \emph{compute\_coverage} & coverage over all ranges \\
  Arithmetic (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
   & \emph{between()} & row-wise gap range \\
   & \emph{span()} & row-wise spanning range \\
   \hline
    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
    & \emph{join\_nearest} & merge by nearest neighbour ranges \\
    & \emph{join\_follow} & merge by following ranges \\
    Merging & \emph{join\_precedes} & merge by preceding ranges \\
    & \emph{union\_ranges} & range-wise union \\
    & \emph{intersect\_ranges} & range-wise intersect \\
    & \emph{setdiff\_ranges} & range-wise set difference \\
    & \emph{complement\_ranges} & range-wise union \\
  \hline
   & \textbf{\emph{anchor\_direction()}} & fix coordinates at direction \\
  Modifier & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
   & \emph{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\emph{filter()}} & subset rows \\
  Restriction & \emph{filter\_by\_overlaps()} & subset by overlap \\
    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
briefly described and categorised into one of: aggregation, unary or binary
arithmetic, merging, modifier, or restriction. A verb is given bold text if
its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
\end{table}

By definition GRanges follows the tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that observation.
Hence, we have designed the \texttt{plyranges} DSL to extend the grammar
and design principles of \texttt{dplyr}: cohesion, consistency,
endomorphism, and fluency. All of these principles are defined and
discussed below in the context of the GRanges class and an overview of
the grammar is provided in table \ref{tab:grammar}. Where applicable we
contrast our design to the existing Bioconductor infrastructure.

\hypertarget{algebraic-operations}{%
\subsection{Algebraic operations}\label{algebraic-operations}}

The \texttt{plyranges} DSL defines an expressive algebra for performing
genomic operations with and between GRanges objects (see table
\ref{tab:grammar}). The grammar includes several classes of operation
that cover most use cases in genomics data analysis. There are operators
for performing arithmetic such as modifying width or performing set
operations, merging such as finding overlaps between GRanges, subsetting
by genomic regions, and aggregating over columns within in a GRanges
object.

GRanges describes the within-sequence coordinates of a range by its
\emph{start}, \emph{end} and \emph{width}. Those three variables are
mutually dependent and partially redundant, so direct manipulation of
them is problematic. For example, changing the \emph{width} column needs
to change either the \emph{start}, \emph{end} or both to preserve
integrity of the object. We introduce the \emph{anchor} modifier to
disambiguate these adjustments. Supported anchor points include the
start, end and midpoint, as well as the 3' and 5' ends for
strand-directed ranges. For example, if we anchor the start, then
setting the width will adjust the end while leaving the start
stationary.

@@

{\centering \includegraphics[width=400pt]{diagrams/diagrams-002}

@@

{\centering \includegraphics[width=400pt]{diagrams/diagrams-003}

@@

Since the GRanges object is a tabular data structure our grammar
includes operators to filter, sort and aggregate by columns in a
GRanges. These operations can be performed over partitions within
GRanges object using the \emph{group\_by} modifier. Together with our
algebra for arithmetic and merging, these operations enable flexible and
reproducible analysis using the desgin principles of the \texttt{dplyr}
grammar.

\hypertarget{cohesion}{%
\subsection{Cohesion}\label{cohesion}}

>>>>

% modified from https://github.com/svmiller/svm-r-markdown-templates
\documentclass[]{article}
\usepackage[top=0.85in,right=2in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{mathpazo}

  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}

\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}

\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center

\renewenvironment{abstract}
 {{%
    \setlength{\leftmargin}{0mm}
    \setlength{\rightmargin}{\leftmargin}%
  }%
  \relax}
 {\endlist}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother

@@

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}

\usepackage{graphicx,grffile}
\usepackage[labelfont=bf]{caption}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}

\title{plyranges: a grammar for transforming genomics data  }

\author{\Large Stuart Lee\vspace{0.05in} \newline\normalsize\emph{Monash University}   \and \Large Michael Lawrence\vspace{0.05in} \newline\normalsize\emph{Genentech}   \and \Large Di Cook\vspace{0.05in} \newline\normalsize\emph{Monash University}  }

\date{}

\usepackage{titlesec}

\titleformat*{\section}{\normalsize\bfseries}
\titleformat*{\subsection}{\normalsize\itshape}
\titleformat*{\subsubsection}{\normalsize\itshape}
\titleformat*{\paragraph}{\normalsize\itshape}
\titleformat*{\subparagraph}{\normalsize\itshape}

% % 
% % \usepackage{biblatex}
% 
\usepackage[backend=biber, style=numeric-comp, url=false, sorting=none]{biblatex}
\addbibresource{references.bib}

\newtheorem{hypothesis}{Hypothesis}
\usepackage{setspace}

\makeatletter
\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \PassOptionsToPackage{hyphens}{url}\usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \PassOptionsToPackage{hyphens}{url}\usepackage[unicode=true]{hyperref}
\fi
}

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Stuart Lee (Monash University) and Michael Lawrence (Genentech) and Di Cook (Monash University)},
             pdfkeywords = {Bioconductor, Grammar, Genomes, Data Analysis},  
            pdftitle={plyranges: a grammar for transforming genomics data},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% add tightlist ----------
\providecommand{\tightlist}{%
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

% \pagenumbering{arabic}% resets `page` counter to 1 
%
% \maketitle

{% \usefont{T1}{pnc}{m}{n}
\setlength{\parindent}{0pt}
\thispagestyle{plain}
{\fontsize{18}{20}\selectfont\raggedright 
\maketitle  % title \par

{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
\textbf{\authorfont Stuart Lee} \hskip 15pt \emph{\small Monash University}   \par \textbf{\authorfont Michael Lawrence} \hskip 15pt \emph{\small Genentech}   \par \textbf{\authorfont Di Cook} \hskip 15pt \emph{\small Monash University}

\begin{abstract}

    \hbox{\vrule height .2pt width 39.14pc}

    \vskip 8.5pt % \small

\noindent The Bioconductor project has created many useful data abstractions for
analysing high-throughput genomics experiments. However, there is a
cognitive load placed on a user in learning a data abstraction and
understanding its appropriate use. Throughout a standard workflow, a
user must navigate and know many of these abstractions to perform an
genomic analysis task, when a single data abstraction, a GRanges object
will suffice. The GRanges class naturally represent genomic intervals
and their associated measurements. By recognising that the GRanges class
follows `tidy' data principles we have created a grammar of genomic data
transformation. The grammar defines verbs for performing actions on and
between genomic interval data. It provides a principled way of
performing common genomic data analysis tasks through a coherent
interface to existing Bioconductor infrastructure, resulting in human
readable analysis workflows. We have implemented this grammar as a
Bioconductor/R package called plyranges.

\vskip 8.5pt \noindent \emph{Keywords}: Bioconductor, Grammar, Genomes, Data Analysis \par

    \hbox{\vrule height .2pt width 39.14pc}

\end{abstract}

\vskip 6.5pt

\noindent  \section{Introduction}\label{introduction}

@@

Several domain specific languages (DSLs) enable scientists to process
and reason about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the Genome Query
Language (GQL) and its distributed implementation GenAp which use an
SQL-like syntax for fast retrieval of information of unprocessed
sequencing data \autocites{Kozanitis2014-va}{Kozanitis2016-bm}.
Similarly, the Genometric Query Language (GMQL) implements a relational
algebra for combining genomic datasets \cite{Kaitoua2017-pw}. The
command line application BEDtools develops an extensive algebra for
performing arithmetic between two or more sets of genomic regions
\cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be
evaluated either at the command line or embedded in scripts for batch
processing. They exist in a sparse ecosystem, mostly consisting of UNIX
and database tools that lack biological semantics and operate at the
level of files and database tables.

The Bioconductor/R packages \texttt{IRanges} and \texttt{GenomicRanges}
\autocites{r-core}{Lawrence2013-wg}{Huber2015-ei} define a DSL for
analysing genomics data with R, an interactive data analysis environment
that encourages reproducibility and provides high-level abstractions for
manipulating, modelling and plotting data, through state of the art
methods in statistical computing. The packages define object-oriented
(OO) abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs that are
embedded in programming languages include pybedtools and valr
\autocites{Dale2011-js}{Kent2017}, however these packages lack the
interoperability provided by the aforementioned Bioconductor packages
and are not easily extended.

@@

Recently, there has been a concerted, community effort to standardize R
data structures and workflows around the notion of tidy data
\cite{Wickham2014-jc}. A tidy dataset is defined as a tabular data
structure that has observations as rows and columns as variables, and
all measurements pertain to a single observational unit. The tidy data
pattern is useful because it allows us to see how the data relate to the
design of an experiment and the variables measured. The \texttt{dplyr}
package \cite{Wickham2017-dplyr} defines an API that maps notions from
the general relational algebra to operations on tidy data. It expresses
each operation as a cohesive, endomorphic verb. Taken together these
features enable a user to write human readable analysis workflows.

@@

\section{Genomic relational algebra}\label{genomic-relational-algebra}

\subsection{Data model}\label{data-model}

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/GRanges}
\caption{An illustration of the GRanges data model for a
sample from an RNA-seq experiment. The core components of the data model
inlcude a seqname column (representing the chromosome), a ranges column
which consists of start and end coordinates for a genomic region, and a
strand identifier (either positive, negative, or unstranded). Metadata
are included as columns to the right of the dotted line as annotations
(gene\_id) or range level covariates (score).}
\label{fig:GRanges} 
\end{figure}

The \texttt{plyranges} DSL is built on the core Bioconductor data
structure GRanges, which is essentially a constrained table, with fixed
columns for the chromosome, start and end coordinates, and the strand,
along with an arbitrary set of additional columns, consisting of
measurements or metadata specific to the data type or experiment (figure
\ref{fig:GRanges}). GRanges balances flexibility with formal
constraints, so that it is applicable to virtually any genomic workflow,
while also being semantically rich enough to support high-level
operations on genomic ranges. As a core data structure, GRanges enables
interoperability between plyranges and the rest of Bioconductor.
Adhering to a single data structure simplifies the API and makes it
easier to learn and understand, in part because operations become
endomorphic, i.e., they return the same type as their input.

GRanges follows the intuitive tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable describing the
observations. GRanges specializes the tidy pattern in that the
observations always pertain to some genomic feature, but it largely
remains compatible with the general relational operations defined by
dplyr. Thus, we define our algebra as an extension of the dplyr algebra,
and borrow its syntax conventions and design principles.

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarise()}} & aggregate over column(s) \\ 
   Aggregation & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighbouring ranges \\
   \hline
   &  \textbf{\emph{mutate()}} & modifies any column \\
   & \textbf{\emph{select()}} & select columns \\
  Arithmetic (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
   & \emph{stretch()} & extend range by fixed amount \\
   &  \emph{shift\_(direction)} & shift coordinates \\
   & \emph{flank\_(direction)} & generate flanking regions \\
   & \emph{\%intersection\% } & row-wise intersection \\
   & \emph{\%union\%} & row-wise union \\
   & \emph{compute\_coverage} & coverage over all ranges \\
  Arithmetic (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
   & \emph{between()} & row-wise gap range \\
   & \emph{span()} & row-wise spanning range \\
   \hline
    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
    & \emph{join\_nearest} & merge by nearest neighbour ranges \\
    & \emph{join\_follow} & merge by following ranges \\
    Merging & \emph{join\_precedes} & merge by preceding ranges \\
    & \emph{union\_ranges} & range-wise union \\
    & \emph{intersect\_ranges} & range-wise intersect \\
    & \emph{setdiff\_ranges} & range-wise set difference \\
    & \emph{complement\_ranges} & range-wise union \\
  \hline
   & \emph{anchor\_direction()} & fix coordinates at direction \\
  Modifier & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
   & \emph{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\emph{filter()}} & subset rows \\
  Restriction & \emph{filter\_by\_overlaps()} & subset by overlap \\
    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
briefly described and categorised into one of: aggregation, unary or binary
arithmetic, merging, modifier, or restriction. A verb is given bold text if
its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
\end{table}

\subsection{Algebraic operations}\label{algebraic-operations}

The \texttt{plyranges} DSL defines an expressive algebra for performing
genomic operations with and between GRanges objects (see table
\ref{tab:grammar}). The grammar includes several classes of operation
that cover most use cases in genomics data analysis. There are range
arithmetic operators, such as for resizing ranges or finding their
intersection, and operators for merging, filtering and aggregating by
range-specific notions like overlap and proximity.

Arithmetic operations transform range coordinates, as defined by their
\emph{start}, \emph{end} and \emph{width}. The three dimensions are
mutually dependent and partially redundant, so direct manipulation of
them is problematic. For example, changing the \emph{width} column needs
to change either the \emph{start}, \emph{end} or both to preserve
integrity of the object. We introduce the \emph{anchor} modifier to
disambiguate these adjustments. Supported anchor points include the
start, end and midpoint, as well as the 3' and 5' ends for
strand-directed ranges. For example, if we anchor the start, then
setting the width will adjust the end while leaving the start
stationary.

@@

{\centering \includegraphics[width=\textwidth]{diagrams/diagrams-002}

@@

{\centering \includegraphics[width=\textwidth]{diagrams/diagrams-003}

@@

Since the GRanges object is a tabular data structure, our grammar
includes operators to filter, sort and aggregate by columns in a
GRanges. These operations can be performed over partitions formed using
the \emph{group\_by} modifier. Together with our algebra for arithmetic
and merging, these operations conform to the semantics and syntax of the
\texttt{dplyr} grammar.

\subsection{Design principles}\label{design-principles}

The design of plyranges adheres to well understood principles of
language and API design: cognitive consistency, cohesion, expressiveness
and endomorphism \cite{Green1996-qg}. To varying degrees, these
principles also underly the design of dplyr and the Bioconductor
infrastructure.

\subsubsection{Cognitive consistency and
fluency}\label{cognitive-consistency-and-fluency}

We aim for our interfaces to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the user
can express workflows in the language of genomics. This motivates the
adoption of the tidy GRanges object as our central data structure. The
basic data.frame and dplyr tibble lack any notion of genomic ranges and
so could not easily support our genomic grammar, with its specific verbs
for range-oriented data manipulation. Another example of cognitive
consistency is how plyranges is insensitive to direction/strand by
default when, e.g., detecting overlaps. GenomicRanges has the opposite
behavior. We believe that defaulting to purely spatial overlap is most
intuitive to most users.
